# Filters and selections

More often than not, we have more data than we want. Sometimes we need to be rid of that data. In `dplyr`, there's two ways to go about this: filtering and selecting.

**Filtering creates a subset of the data based on criteria**. All records where the count is greater than 10. All records that match "Nebraska". Something like that. **Filtering works with rows -- when we filter, we get fewer rows back than we start with.**

**Selecting simply returns only the fields named**. So if you only want to see Year and County, you select those fields. When you look at your data again, you'll have two columns. If you try to use one of your columns that you had before you used select, you'll get an error. **Selecting works with columns. You will have the same number of records when you are done, but fewer columns of data to work with.**

Let's work with the [salaries data from the University of Nebraska](https://unl.box.com/s/9826nisk29fztlc1xhup988eah0mqdby). It has data from all NU campuses, but only one of them is our campus, so let's filter out everyone else. 

```{r}
library(tidyverse)
```

```{r}
salaries <- read_csv("data/nusalaries1819.csv")
```

The data we want to filter on is in `Campus`. So we're going to use filter and something called a comparison operator. We need to filter all records equal to UNL. The comparison operators in R, like most programming languages, are == for equal to, != for not equal to, > for greater than, >= for greater than or equal to and so on. 

**Be careful: `=` is not `==` and `=` is not "equal to". `=` is an assignment operator in most languages -- how things get named.** 

```{r}
unl <- salaries %>% filter(Campus == "UNL")

head(unl)
```

And just like that, we have just UNL, which we can verify looking at the head, the first six rows. 

We also have more data than we might want. For example, the salary data is only in the Budgeted Annual Salary column. The other two salary fields are useless detail. 

To simplify our dataset, we can use select. 

```{r}
selected_unl <- unl %>% select(Employee, Position, Campus, `Budgeted Annual Salary`)

head(selected_unl)
```

And now we only have four columns of data for whatever salary analysis we might want to do. 
