---
title: "lab_06"
author: "derek willis"
date: "8/26/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## You will need

* Tabula
* The [Atom](https://atom.io/) text editor.

## Load libraries and establish settings
```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse.
library(tidyverse)
```

## Get Our PDF

We'll be working with the [Maryland Statistical Handbook](https://planning.maryland.gov/MSDC/Documents/md-statistical-handbook.pdf) from the state Department of Planning. It's not tiny (44MB), and you'll want to download it to a place you'll remember (like your Downloads folder).

## Setup Tabula

Start Tabula, then go to http://127.0.0.1:8080/ in your browser. Click the "Browse" button and find the Statistical Handbook file and click "open", and then click the "Import button" in Tabula. This will take a minute or two.

Let's go to page 30 of the PDF, Table 2A, "International Migration for Maryland's Jurisdictions, July 1, 2010 to July 1, 2019". This is _net migration_, meaning it shows the total number of people moving to Maryland and its regions/jurisdictions _minus_ the total number of people leaving the same, divided into 12-month periods from July to July. In Tabula, draw a box around that table's border and click the "Preview & Export Extracted Data" button. It should look pretty clean. Let's export that CSV (it should be called `tabula-md-statistical-handbook.csv` by default) to your Downloads folder.

## Cleaning up the data in R

Let's load it into R, and in doing so we'll skip the first two rows and add our own headers that are cleaner. `read_csv` allows us to do this ([and more](https://readr.tidyverse.org/reference/read_delim.html)):

```{r}
international_migration <- read_csv('tabula-md-statistical-handbook.csv', skip=2, col_names=c("jurisdiction", "july_2011", "july_2012", "july_2013", "july_2014", "july_2015", "july_2016", "july_2017", "july_2018", "july_2019", "total"))
```

Add a column for the type of migration ("international") and populate it:

```{r}
international_migration <- international_migration %>% mutate(type='international')
```

## Answer questions

Q1. Which region and county/city below the state level accounted for the largest percentage of international migration overall?  You'll need to add and populate columns representing percent of total using `mutate`.

Q2. Write a sentence or two that describes the data you produced in A1. Treat this as if you were trying to convey the most important part.

Q3. Which region & jurisdiction had the biggest percentage change for international migration between July 2018 and July 2017? The formula for percentage change is easy to remember: (New-Old)/Old.

Q4. What's your best guess as to why these declines occurred, and in those area in particular?

## Back to Tabula

Let's go to page 31 of the PDF, Table 2B, "Domestic Migration for Maryland's Jurisdictions, July 1, 2010 to July 1, 2019". In Tabula, hit the "Clear All Selections" button and then draw a box around that table's border and click the "Preview & Export Extracted Data" button. It should look pretty clean. Let's export that CSV to your Downloads folder (let's rename it to `tabula-md-statistical-handbook-domestic.csv`).

## Cleaning up the data in R

Let's load it into R, and in doing so we'll skip the first two rows and add our own headers that are cleaner:

```{r}
domestic_migration <- read_csv('tabula-md-statistical-handbook-domestic.csv', skip=2, col_names=c("jurisdiction", "july_2011", "july_2012", "july_2013", "july_2014", "july_2015", "july_2016", "july_2017", "july_2018", "july_2019", "total"))
```

Add a column for the type of migration ("domestic") and populate it:

```{r}
domestic_migration <- domestic_migration %>% mutate(type='domestic')
```

## Answer questions
Q5. Which Maryland individual jurisdiction saw the largest net decrease in domestic migration overall?

Q6. How many regions & jurisdictions had net positive migration for July 2017, July 2018 and July 2019?

Q7. How would you describe this data? Is there a county or region that stands out, and why?
