---
title: "lab_01"
author: "sean mussenden"
date: "8/19/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# Turn off scientific notation
options(scipen=999)
```

## About this lab

To complete this lab, you need to:
* run existing code as directed (look for **Task**). 
* modify existing code as directed (look for **Task**).
* write code in empty codeblocks provided to answer questions included (look for **Q**). 
* write out the answer in the form of a complete sentence in the space given (look for **A**). 
When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS. 

## Load libraries and establish settings

**Task** Run the codeblock below to load the Tidyverse. To run the codeblock, hit the little green play button in the upper right corner of the gray area (the codeblock) below.

```{r}

# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
```

## Load Data 

We're going to load the slice of PPP loans for West Virginia.

**Task** Run the codeblock below to load the data. 

```{r}
wv_ppp_loans <- read_rds("lab_01.rds")
```

## Examine Data

**Task** Run the codeblock below to use glimpse to get a sense of the column names, types and sample data. 
**Q1** How many rows and columns does this dataframe have?
**A1** 33,703 rows, 63 columns.

```{r}

glimpse(wv_ppp_loans)

```

## Answer questions

**Q2** Which West Virginia county got the most loans? How many?
**A2** Kanawha County, West Virginia got 4,332 loans, more than any other West Virginia county.  

```{r}
wv_ppp_loans %>%
  group_by(project_county_name) %>%
  summarise(
    total_loans = n()
  ) %>%
  arrange(desc(total_loans))
```
**Q3** Which West Virginia county got the fewest loans? How many?
**A3** Wirt County, West Virginia got the feweest loans of any West Virginia county, with 36 loans. 
```{r}
wv_ppp_loans %>%
  group_by(project_county_name) %>%
  summarise(
    total_loans = n()
  ) %>%
  arrange(total_loans)
```
**Q4** What's your best educated guess to explain why the county identified in question 2 had a lot of loans and why the county identified in question 3 had only a few loans? Hint: you won't find the answer in the data, you'll need to do some quick web research. You do not need to write any code for this question, so no codeblock is provided.  
**A4** One possible explanation is that a lot of people live in Kanawha County, and few people live in Wirt County. According to 2019 U.S. Census American Community Survey 5-year population estimates, Kanawha is West Virginia's largest county with 183,279 people, and Wirt is its smallest, with 5,798 people. It's reasonable to expect a place with a lot more people to have a lot more loans. It's often important to account for population size when drawing conclusions from data analysis. [census source]([source](https://data.census.gov/cedsci/table?q=United%20States&g=0400000US54.0500000&tid=ACSDT5Y2019.B01003&hidePreview=true&tp=true&moe=false) 


**Q5** What West Virginia business got the state's largest PPP loan? How much was it for? What does the business do? How many jobs did the loan retain, according to the data?
**A5** The Greenbriar Hotel Corporation got a loan of $8,912,202, more than any other business in the state. In a harbinger of things to come when working with PPP loan data, the name of the company is misspelled in the data. The correct name is the Greenbrier Hotel Corporation.  The company runs The Greenbrier, an 11,000 acre luxury resort in White Sulphur Springs that houses a secret bomb shelter that, at one time, was designed to protect Congress in the event of nuclear war. According to the PPP loan data, it retained 500 jobs, though ProPublica reported that the company originally listed 0 as the number of jobs that would be retained. 

Links: 
* [The Greenbrier](https://www.greenbrier.com/about-us.aspx)
* [the-secret-bunker-congress-never-used](https://www.npr.org/2011/03/26/134379296/the-secret-bunker-congress-never-used) 

```{r}
wv_ppp_loans %>%
  arrange(desc(amount))
```

**Q6** What West Virginia politician (with his family) owns the company identified in question 2? What's your source for that information? Hint: this will require web research, you won't be able to answer this purely from the data set.  You do not need to write any code for this question, so no codeblock is provided.  
**A6** West Virginia Gov. Jim Justice -- a billionaire -- and his family own the Greenbrier. Justice has been the frequent subject of investigative stories detailing the many conflicts of interests between his role as governor and his businesses.  According to a ProPublica story from July 2020, his companies got up to $24 million in bailout loans. 

Links: 
* [Companies Owned by This Billionaire Governor Received up to $24 Million in Bailout Loans](https://www.propublica.org/article/companies-owned-by-this-billionaire-governor-received-up-to-24-million-in-bailout-loans)
* [Big Jim: West Virginia’s Conflicted Governor](https://www.propublica.org/series/big-jim)


**Q7**. Produce a summary table that has the following information for West Virginia as a whole (the whole state, not by county):
* count of loans
* total amount of all loans
* the average loan amount
* the median loan amount
* the smallest loan amount
* the largest loan amount

Store it as a variable object called `wv_summary` so that you can examine it with the data viewer. What newsworthy nuggets jump out at you that you could imagine dropping into a news story? Write a few summary sentences with overall findings. 

**A7**. Nearly 34,000 businesses in West Virginia got a total of \$2.6 billion in Paycheck Protection Program loans since the launch of the program. The average loan was \$77,338, but the median loan amount was only \$20,832, indicating several massive loans are skewing the average upward. The largest loan in the state was \$8.9 million, but the smallest was only $165.  It might be worth examining the small loans to identify patterns there. 

```{r}
wv_summary <- wv_ppp_loans %>%
  summarise(
    count_loans = n(),
    total_loans_amount = sum(amount),
    mean_loan_amount = mean(amount),
    median_loan_amount = median(amount),
    min_loan_amount = min(amount),
    max_loan_amount = max(amount)
  ) %>% 
  arrange(desc(max_loan_amount))
```

**Q8** Produce a summary table that has the following information for each West Virginia county:
* count of loans
* total amount of all loans
* the average loan amount
* the median loan amount
* the smallest loan amount
* the largest loan amount

Store it as a variable object called `wv_county_summary` so that you can examine it with the data viewer. 

What newsworthy nuggets or patterns jump out at you that you could imagine dropping into a news story or doing follow up reporting on? Write up a few sentence that summarizes one county. 

**A8** 
```{r}
wv_county_summary <- wv_ppp_loans %>%
  group_by(project_county_name) %>%
  summarise(
    count_loans = n(),
    total_loans_amount = sum(amount),
    mean_loan_amount = mean(amount),
    median_loan_amount = median(amount),
    min_loan_amount = min(amount),
    max_loan_amount = max(amount)
  ) %>% 
  arrange(desc(max_loan_amount))
```
